Supplementary Materials And Methods

### 2019-09-29 NAR_form_article.Rmd
scanning

## Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns. After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly. Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

## Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (Figure \ref{fig:fig1}A) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (Figure \ref{fig:fig1}B). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (Figure \ref{fig:fig1}D), consequently have potential to affect whole network due to "ripple effect".

@@

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in Figure \ref{fig:fig2}E.

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. SERPINE2, one of the most effective nodes, has been used as trigger in ceRNA network simulation in breast cancer patient network. Relevant steps and codes are available at  Supplementary Materials and Methods.

@@

## Network with miRNA:ceRNA expression level
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\ref{fig:fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, changes in expression levels of G4 and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (Figure \ref{fig:fig1}D).

@@

## ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in Figure \ref{fig:fig1} is recalculated with additional factors Table \ref{tab:one} in effect with same trigger, two fold increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in Figure \ref{fig:fig3}A. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.

@@

It is considered that entire miRNAs in the system affect targets according to target:total target ratio Figure \ref{fig:fig1}A, when interaction factors did not take into account. However, in presence of interaction factors in miRNA:target interaction network (shown in Table \ref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in Figure \ref{fig:fig2}C. After affinity mediated proportional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in Figure \ref{fig:fig2}D). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 Figure \ref{fig:fig3}A in comparison with miRNA1 (M1):Gene2 (G2) interaction in Figure \ref{fig:fig1}A. On the other hand, it is possible that taking into account of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction Figure \ref{fig:fig3}A. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (Figure \ref{fig:fig3} B,C,D). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. *(see details in Supplementary Materials and Methods*. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

@@

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state *(see details in Supplementary Materials and Methods)*. Simulations show that change in expression level of single gene has potential to affect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

We have observed that the results of simulation with SERPINE2 gene, a serine protease inhibitor, one of the overexpressed genes of breast cancer [@yang2018expression; @candia2006protease] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach. Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.

>>>>

## Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns (see Figure S1 in Supplementary Tables). After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly. Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

## Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (Figure \ref{fig:fig1}A) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (Figure \ref{fig:fig1}B). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (Figure \ref{fig:fig1}D), consequently have potential to affect whole network due to "ripple effect" (see Section 2 in Supplementary Materials and Methods).

@@

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in Figure \ref{fig:fig2}E (see Section 3 in Supplementary Materials and Methods).

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015] (Detailed codes are available at Section 6 in Supplementary Materials and Methods). We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. SERPINE2, one of the most effective nodes, has been used as trigger in ceRNA network simulation in breast cancer patient network. Relevant steps and codes are available at Supplementary Materials and Methods.

@@

## Network with miRNA:ceRNA expression level
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\ref{fig:fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression.

@@

## ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in Figure \ref{fig:fig1} is recalculated with additional factors Table \ref{tab:one} in effect with same trigger, two fold increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in Figure \ref{fig:fig3}A. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.  On the other hand, changes in expression levels of G4 and G1 differ although they have same initial conditions (Detailed codes are available at Sections 3-4 in Supplementary Materials and Methods). This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (Figures \ref{fig:fig3}D) and changed interaction factors.

@@

It is considered that entire miRNAs in the system affect targets according to target:total target ratio Figure \ref{fig:fig1}A, when interaction factors were not taken into account. However, in presence of interaction factors in miRNA:target interaction network (shown in Table \ref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in Figure \ref{fig:fig2}C. After affinity mediated proportional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in Figure \ref{fig:fig2}D). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 Figure \ref{fig:fig3}A in comparison with miRNA1 (M1):Gene2 (G2) interaction in Figure \ref{fig:fig1}A. On the other hand, it is possible that taking into account of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction Figure \ref{fig:fig3}A. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (Figure \ref{fig:fig3} B,C,D). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies (Detailed codes are available at Section 4 in Supplementary Materials and Methods). On the other hand, when we applied the method on the minimal dataset (i.e. Gene4 is trigger element), Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

@@

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state (see details at Section 6 in Supplementary Materials and Methods). Simulations show that change in expression level of single gene has potential to affect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

We have analysed the realistic network in context of perturbation efficiencies of nodes. This process were run with 3 fold upregulation and 10 iteration in conditions where elements that have less than five percent change are not taken into account (Codes are found at Section 6.5 in Supplementary Materials and methods). With this approach, SERPINE is found as one of the most effective gene nodes in network. We have observed that the results of simulation with SERPINE2 gene, a serine protease inhibitor, one of the overexpressed genes of breast cancer [@yang2018expression; @candia2006protease] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach (see Section 6.6 in Supplementary Materials and methods). Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.

### 2019-09-29 NAR_form_article.tex
scanning

miRNA and target pairs per line should be provided as edge list to
construct the network. At each line minimum required information is
expression levels of miRNA and the target. If available, additional data
about factors effecting binding or efficiency of miRNA can be provided
as separate columns. After construction of the network, amount of miRNA
per target is calculated and kept as edge data. Simply, a target will
sequester miRNA proportional to its ratio amount among other targets. If
additional criteria effecting the binding of miRNA to its target is
provided, distribution of miRNA will be calculated accordingly. Target
can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout
the manuscript terms target, gene and ceRNA are used interchangeably.

@@

Initially, the network is assumed in steady-state (Figure
\ref{fig:fig1}A) condition and needs least one trigger for initiating
calculations. The trigger can be a change in expression level of one or
more genes (Figure \ref{fig:fig1}B). After a trigger, the network
undergoes iterative cycle of calculations at each of which distribution
of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C).
Based on new miRNA distribution, expression level of each node
(i.e.~ceRNA) is updated. This update results in change of expression
value of common target (G4) in system. In this case, the common target
acts as trigger for the other group (the targets interacted with M2
miRNA) in the network. Due to common targeted elements, the change in
one neighborhood spreads to other neighborhoods (Figure
\ref{fig:fig1}D), consequently have potential to affect whole network
due to ``ripple effect''.

@@

Not all miRNA:target binding events result in degradation of target. The
degradation of target by bound miRNA depends on efficiency factors such
as binding region. Exact repression efficiency of miRNA is calculated
according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D);
\(RE^\prime_{gi}\), normalized values of region efficiency coefficient
between miRNA and gene. The cooperative repression activity of miRNAs to
their common targets is figured out as shown in Figure \ref{fig:fig2}E.

@@

We have applied our model in a real dataset for which experimental
measurements of various factors were available. Expression levels of
miRNA and genes in tumor and normal tissue of single patient are
retrieved from TCGA Research Network \url{https://www.cancer.gov/tcga}.
High-throughput experimental datasets which are provided miRNA:gene
target pairs with interaction factors
\citep{helwak_mapping_2013, moore_mirnatarget_2015}. We have combined
miRNA and gene expression datasets via miRNA:target gene dataset
retrieved from published datasets
\citep{helwak_mapping_2013, moore_mirnatarget_2015}. Detailed
description of network construction and its code is available in
Supplementary Materials and Methods. SERPINE2, one of the most effective
nodes, has been used as trigger in ceRNA network simulation in breast
cancer patient network. Relevant steps and codes are available at
Supplementary Materials and Methods.

@@

We have developed a network-based approach to assess effects of
expression level changes in competitive ceRNA regulation. The basic mode
of miRNA repression activity has been based on miRNA and target
abundance in various researches
\citep{arvey_target_2010, denzler_assessing_2014}. Our approach can
effortlessly calculate effects of expression changes when abundance
levels of miRNAs and targets is only available factor. In sample network
given in Figure (\ref{fig:fig1}), after an increase in expression level
of a gene (G2), expression values of other genes also changed due to
redistribution of miRNA among its targets. Previous studies have shown
that if a gene abundance increases in ceRNA system, expression levels of
genes targeted by shared miRNA are also affected
\citep{lai_understanding_2016, salmena_cerna_2011, tay_multilayered_2014}.
It was observed that expression levels of primary neighborhoods which do
not interact with another miRNA of the trigger gene change in relation
to their expression. However, changes in expression levels of G4 and G1
differ although they have same initial conditions. This is due to fact
that G4 is targeted by two miRNAs, M1 and M2 (Figure \ref{fig:fig1}D).

@@

Earlier studies reported that miRNA regulatory interactions are affected
by different parameters. For example, Xu et al.~have investigated the
importance of seed pairing type between miRNAs and their targets and
target site location by using proteomics dataset
\citep{xu_characterization_2014}. They have proposed that the features
of binding between miRNA and target can be critical for miRNA
efficiency. In addition, binding energy between miRNA and targets is a
significant determinant for miRNA efficiency and it has been reported
that strength of miRNA:target interactions is depended on binding energy
of complexes \citep{breda_quantifying_2015}. Similarly, another study
revealed that affinity is correlated with seed pairing of miRNA:target
pairs and suggested that affinity is correlated with number of canonical
seed base pairing \citep{bosson_endogenous_2014}. Therefore, we have
integrated aforementioned interaction parameters that could be useful
for predicting miRNA repression activity more accurately. The sample
dataset used in Figure \ref{fig:fig1} is recalculated with additional
factors Table \ref{tab:one} in effect with same trigger, two fold
increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken
into account in the system, miRNA efficiencies varied as shown in Figure
\ref{fig:fig3}A. Although the miRNA:target expression ratios in
steady-state were same in comparison with the sample dataset without
factors, efficiency of binding and repression have changed.

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3}),
miRNA (M1) repressive efficiency on the primary triggering gene (G2) is
low in steady-state. So it has been observed that the regulatory
activities of miRNA on the targets are weak after the increase of miRNA
target G2. When model is triggered with two fold increase in expression
level of common target (G4), more prominent changes were observed in
gene expression levels in network compared to changes caused by G2
increase. Furthermore, change in expression level of target gene that
has strong miRNA repression efficiency resulted in evident perturbation
in network. On the other hand, it was observed that Gene2 was weakly
affected from change of Gene4 expression because of its weak interaction
factors. Since perturbation efficiency of each gene is different, we
have developed a function which screens each gene in the network for
their perturbation efficiencies. \emph{(see details in Supplementary
Materials and Methods}. On the other hand, when we applied the method on
the minimal dataset, Gene4 has been found to be the most efficient
element in terms of number of perturbed elements and miRNA2 (M2) has
been found to be causing the highest mean in expression changes.

@@

The network was finalized by integrating gene and miRNA expression
levels from TCGA determined by RNA-Seq. Our model can successfully
simulate perturbations in such large network despite complex behaviors
and struggle to reach steady-state \emph{(see details in Supplementary
Materials and Methods)}. Simulations show that change in expression
level of single gene has potential to affect whole network, perturbing
almost all nodes. These observations are in accordance with competing
endogenous RNA hypothesis where genes targeted with many common miRNAs
subsequently transmit perturbation to neighboring groups.

We have observed that the results of simulation with SERPINE2 gene, a
serine protease inhibitor, one of the overexpressed genes of breast
cancer \citep{yang2018expression, candia2006protease} is not convenient
with tumor tissue expression values, when we compared the tumor gene
expression dataset retrieved from TCGA and simulation results obtained
via our approach. Simulation results were not expected to be consistent
with tumor tissue expression values because large number of abnormal
regulations were observed in these tissues. So, it might not said that a
single gene is responsible for entire regulations in cancer cells.
Besides, the all other factors such as up/down regulation of miRNAs or
other genes were ignored. Therefore, it would be more useful to test our
approach with an ideal dataset that includes entire informations of gene
expression values before and after the regulation of a known gene at
existing tissue, and miRNA expression values of this tissue at initial
conditions.

>>>>

miRNA and target pairs per line should be provided as edge list to
construct the network. At each line minimum required information is
expression levels of miRNA and the target. If available, additional data
about factors effecting binding or efficiency of miRNA can be provided
as separate columns (see Figure S1 in Supplementary Tables). After
construction of the network, amount of miRNA per target is calculated
and kept as edge data. Simply, a target will sequester miRNA
proportional to its ratio amount among other targets. If additional
criteria effecting the binding of miRNA to its target is provided,
distribution of miRNA will be calculated accordingly. Target can be mRNA
or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the
manuscript terms target, gene and ceRNA are used interchangeably.

@@

Initially, the network is assumed in steady-state (Figure
\ref{fig:fig1}A) condition and needs least one trigger for initiating
calculations. The trigger can be a change in expression level of one or
more genes (Figure \ref{fig:fig1}B). After a trigger, the network
undergoes iterative cycle of calculations at each of which distribution
of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C).
Based on new miRNA distribution, expression level of each node
(i.e.~ceRNA) is updated. This update results in change of expression
value of common target (G4) in system. In this case, the common target
acts as trigger for the other group (the targets interacted with M2
miRNA) in the network. Due to common targeted elements, the change in
one neighborhood spreads to other neighborhoods (Figure
\ref{fig:fig1}D), consequently have potential to affect whole network
due to ``ripple effect'' (see Section 2 in Supplementary Materials and
Methods).

@@

Not all miRNA:target binding events result in degradation of target. The
degradation of target by bound miRNA depends on efficiency factors such
as binding region. Exact repression efficiency of miRNA is calculated
according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D);
\(RE^\prime_{gi}\), normalized values of region efficiency coefficient
between miRNA and gene. The cooperative repression activity of miRNAs to
their common targets is figured out as shown in Figure \ref{fig:fig2}E
(see Section 3 in Supplementary Materials and Methods).

@@

We have applied our model in a real dataset for which experimental
measurements of various factors were available. Expression levels of
miRNA and genes in tumor and normal tissue of single patient are
retrieved from TCGA Research Network \url{https://www.cancer.gov/tcga}.
High-throughput experimental datasets which are provided miRNA:gene
target pairs with interaction factors
\citep{helwak_mapping_2013, moore_mirnatarget_2015} (Detailed codes are
available at Section 6 in Supplementary Materials and Methods). We have
combined miRNA and gene expression datasets via miRNA:target gene
dataset retrieved from published datasets
\citep{helwak_mapping_2013, moore_mirnatarget_2015}. Detailed
description of network construction and its code is available in
Supplementary Materials and Methods. SERPINE2, one of the most effective
nodes, has been used as trigger in ceRNA network simulation in breast
cancer patient network. Relevant steps and codes are available at
Supplementary Materials and Methods.

@@

We have developed a network-based approach to assess effects of
expression level changes in competitive ceRNA regulation. The basic mode
of miRNA repression activity has been based on miRNA and target
abundance in various researches
\citep{arvey_target_2010, denzler_assessing_2014}. Our approach can
effortlessly calculate effects of expression changes when abundance
levels of miRNAs and targets is only available factor. In sample network
given in Figure (\ref{fig:fig1}), after an increase in expression level
of a gene (G2), expression values of other genes also changed due to
redistribution of miRNA among its targets. Previous studies have shown
that if a gene abundance increases in ceRNA system, expression levels of
genes targeted by shared miRNA are also affected
\citep{lai_understanding_2016, salmena_cerna_2011, tay_multilayered_2014}.
It was observed that expression levels of primary neighborhoods which do
not interact with another miRNA of the trigger gene change in relation
to their expression.

@@

Earlier studies reported that miRNA regulatory interactions are affected
by different parameters. For example, Xu et al.~have investigated the
importance of seed pairing type between miRNAs and their targets and
target site location by using proteomics dataset
\citep{xu_characterization_2014}. They have proposed that the features
of binding between miRNA and target can be critical for miRNA
efficiency. In addition, binding energy between miRNA and targets is a
significant determinant for miRNA efficiency and it has been reported
that strength of miRNA:target interactions is depended on binding energy
of complexes \citep{breda_quantifying_2015}. Similarly, another study
revealed that affinity is correlated with seed pairing of miRNA:target
pairs and suggested that affinity is correlated with number of canonical
seed base pairing \citep{bosson_endogenous_2014}. Therefore, we have
integrated aforementioned interaction parameters that could be useful
for predicting miRNA repression activity more accurately. The sample
dataset used in Figure \ref{fig:fig1} is recalculated with additional
factors Table \ref{tab:one} in effect with same trigger, two fold
increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken
into account in the system, miRNA efficiencies varied as shown in Figure
\ref{fig:fig3}A. Although the miRNA:target expression ratios in
steady-state were same in comparison with the sample dataset without
factors, efficiency of binding and repression have changed. On the other
hand, changes in expression levels of G4 and G1 differ although they
have same initial conditions (Detailed codes are available at Sections
3-4 in Supplementary Materials and Methods). This is due to fact that G4
is targeted by two miRNAs, M1 and M2 (Figures \ref{fig:fig3}D) and
changed interaction factors.

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3}),
miRNA (M1) repressive efficiency on the primary triggering gene (G2) is
low in steady-state. So it has been observed that the regulatory
activities of miRNA on the targets are weak after the increase of miRNA
target G2. When model is triggered with two fold increase in expression
level of common target (G4), more prominent changes were observed in
gene expression levels in network compared to changes caused by G2
increase. Furthermore, change in expression level of target gene that
has strong miRNA repression efficiency resulted in evident perturbation
in network. On the other hand, it was observed that Gene2 was weakly
affected from change of Gene4 expression because of its weak interaction
factors. Since perturbation efficiency of each gene is different, we
have developed a function which screens each gene in the network for
their perturbation efficiencies (Detailed codes are available at Section
4 in Supplementary Materials and Methods). On the other hand, when we
applied the method on the minimal dataset (i.e.~Gene4 is trigger
element), Gene4 has been found to be the most efficient element in terms
of number of perturbed elements and miRNA2 (M2) has been found to be
causing the highest mean in expression changes.

@@

The network was finalized by integrating gene and miRNA expression
levels from TCGA determined by RNA-Seq. Our model can successfully
simulate perturbations in such large network despite complex behaviors
and struggle to reach steady-state (see details at Section 6 in
Supplementary Materials and Methods). Simulations show that change in
expression level of single gene has potential to affect whole network,
perturbing almost all nodes. These observations are in accordance with
competing endogenous RNA hypothesis where genes targeted with many
common miRNAs subsequently transmit perturbation to neighboring groups.

We have analysed the realistic network in context of perturbation
efficiencies of nodes. This process were run with 3 fold upregulation
and 10 iteration in conditions where elements that have less than five
percent change are not taken into account (Codes are found at Section
6.5 in Supplementary Materials and methods). With this approach, SERPINE
is found as one of the most effective gene nodes in network. We have
observed that the results of simulation with SERPINE2 gene, a serine
protease inhibitor, one of the overexpressed genes of breast cancer
\citep{yang2018expression, candia2006protease} is not convenient with
tumor tissue expression values, when we compared the tumor gene
expression dataset retrieved from TCGA and simulation results obtained
via our approach (see Section 6.6 in Supplementary Materials and
methods). Simulation results were not expected to be consistent with
tumor tissue expression values because large number of abnormal
regulations were observed in these tissues. So, it might not said that a
single gene is responsible for entire regulations in cancer cells.
Besides, the all other factors such as up/down regulation of miRNAs or
other genes were ignored. Therefore, it would be more useful to test our
approach with an ideal dataset that includes entire informations of gene
expression values before and after the regulation of a known gene at
existing tissue, and miRNA expression values of this tissue at initial
conditions.

### 2019-09-29 Supplementary_mat_met.Rmd
scanning

## *minsamp* dataset analysis in lack of interaction factors.

@@

## *minsamp* dataset analysis with interaction factors.

@@

## Common target perturbation in *minsamp* dataset.

Genes targeted by multiple miRNAs (referred to as "common target") are of special interest since they are subject to cooperative effect. Also, they perturb more than one neighborhood. In our small dataset, minsamp, Gene4 is regulated by two miRNAs. Let's simulate perturbation effects triggered by Gene4 (Shown at Figure S8 in Supplementary Figures .

@@

## Determination of perturbation efficiencies of elements in system.

@@

# 2. Obtaining breast cancer dataset and integration

@@

## 2.1 How to get gene expression counts of TCGA-E9-A1N5 patient.

@@

## 2.2 How to get miRNA expression counts of TCGA-E9-A1N5 patient.

@@

## 2.3 Get the high-throughput experimental miRNA:target dataset.

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We committed the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values at Supplementary files, Process on miRNA:target pairs dataset for process codes and Supplementary Tables for used numeric values)

@@

## 2.4 Combine the dataset

@@

## 2.5 Selection of trigger node

@@

## 2.5 Approach of Method into Combined Datasets

@@

### 2.5.1 Find iteration of simulation

@@

### 2.5.2 Simulation of dataset

@@

### 2.5.3 Comparison of simulation results and tumor tissue expression values
```{r}
E9GE_mirnagenetumor%>%
  dplyr::select(hgnc_symbol, GE_tumor)%>%
  dplyr::inner_join((E9GE_mirnagenenormal%>%dplyr::select(hgnc_symbol, GE_normal)), by ="hgnc_symbol")%>%
  inner_join(simulation_results, by= c("hgnc_symbol"="name"))%>%
  distinct()
```

>>>>

# 2. *minsamp* dataset analysis in lack of interaction factors.

@@

# 3. *minsamp* dataset analysis with interaction factors.

@@

# 4. Common target perturbation in *minsamp* dataset.

Genes targeted by multiple miRNAs (referred to as "common target") are of special interest since they are subject to cooperative effect. Also, they perturb more than one neighborhood. In our small dataset, minsamp, Gene4 is regulated by two miRNAs. Let's simulate perturbation effects triggered by Gene4 (Shown at Figure S8 in Supplementary Figures) .

@@

# 5. Determination of perturbation efficiencies of elements in system.

@@

# 6. Obtaining breast cancer dataset and integration

@@

## 6.1 How to get gene expression counts of TCGA-E9-A1N5 patient.

@@

## 6.2 How to get miRNA expression counts of TCGA-E9-A1N5 patient.

@@

## 6.3 Get the high-throughput experimental miRNA:target dataset.

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We committed the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values. The codes about integration of high-throughput experimental studies are available at Supplementary additional file and numeric values of interaction factors can be found at Table S2-3 in Supplementary Tables file.)

@@

## 6.4 Combine the dataset

@@

## 6.5 Selection of trigger node

@@

## 6.6 Approach of Method into Combined Datasets

@@

### 6.6.1 Find iteration of simulation

@@

### 6.6.2 Simulation of dataset

@@

### 6.6.3 Comparison of simulation results and tumor tissue expression values
```{r}
E9GE_mirnagenetumor%>%
  dplyr::select(hgnc_symbol, GE_tumor)%>%
  dplyr::inner_join((E9GE_mirnagenenormal%>%dplyr::select(hgnc_symbol, GE_normal)), by ="hgnc_symbol")%>%
  inner_join(simulation_results, by= c("hgnc_symbol"="name"))%>%
  distinct()
```
